# Using loops to calculate sums and sums of squares
Please copy-paste code from the previous exercise
over to a new code block 

## Printing values {#printEx}
Here we have the following list of values:
```{r eval=FALSE}
my.list <- c(5,10,19,22,3,40,48)
```
Produce a for loop that uses the `print()` function
to print each value to the console, e.g., 
```
[1] 5
[1] 10
[1] 19
[1] 22
[1] 3
[1] 40
[1] 48
```


## A simple sum {#sum}
Use a single `for` loop to calculate the sum of `my.list`
in exercise \@ref(printEx). This is chiefly for the sake of 
practice as in reality we would use the `sum()`
function for this.

## Sums of squares {#sumSq}
Now copy-paste and modify the code from exercise \@ref(sum)
to calculate both the sum
and the sum of squares within one and the same `for`-loop.
Calculating the sum of squares is the same as first squaring each 
numbers and then summing these squares, i.e., $5\times 5 + 10 \times 10 + \cdots + 48 \times 48$.

The program should have no output, other than the
following message at the end: 

`[1] The sum is 147 and the sum of squares is 4883`

which can be done with the functions [`print()`](  https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/print)
and, in particular, [`paste()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/paste)

## Sum only if larger than $x$
Copy-paste the code from section \@ref(sumSq)
and use an `if`-statement 
to calculate the sum and 
the sum of squares for those numbers in the list above
that are greater or equal than $x=10$. Other numbers
can be ignored. The resulting message should now be 

`[1] The sum is 139 and the sum of squares is 4849`


## Separate sums for odd and even numbers
Copy-paste the code from exercise \@ref(sumSq) 
and modify it to calculate sums and 
sums of squares 
for odd and even numbers separately. 
Do this using an `if`-`else`
statement. The resulting code should print
the following message:
```
[1] "The sum of even numbers is 120 and their sum of squares is 4488. The sum of odd numbers is 30 and their sum of squares is 395."
```
Note the punctuation in the output above, which was
not there before and requires some changes in the `paste()`
statement. 

Hint: to find out whether a number is odd or even,
use the modulo operator `a %% b`. This gives the remainder of
the division `a/b`. For example,
```{r eval=FALSE}
33 %% 3 #0
33 %% 2 # 1
33 %% 5  # 3
33 %% 10  # 3
```
