# Slightly more advanced for loops
In this set of exercises we will look at slightly more complicated for loops. 

```{r include=FALSE}
# https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```

## Exercise: select columns of a `data.frame` containing floating-point numbers
Data types like [`data.frames`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) and [`tibbles`](https://tibble.tidyverse.org/) are collections of variables ordered in rows and columns, like a spreadsheet. Take for example the `Cars93` `data.frame` from R's `MASS` package, which we inspect with the [`str()`](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/str.html) command:
```{r collapse=T, output.lines=6}
library("MASS")
str(Cars93)
```

Sometimes, we would like to select certain columns from such a `data.frame`, dependent on a particular condition. Hence the current question:

**Return a data.frame in which the columns contain floating-point numbers, such as `6.5` or `1e-03`. Omit any other column. Use a `for`-loop rather than tidyverse commands such as `filter()` or `select()`.**

### Hint 1: data types
The way to find out whether a column of a `data.frame` contains floating-point numbers is to use the [`typeof()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/typeof.html) function. Let us compare two different columns from the `Cars93` dataset, for example: 
```{r, collapse=T}
typeof(Cars93[,"Min.Price"])
typeof(Cars93[,"MPG.city"])
```

where `double` reflects that the `Min.Price` column contains floating-point numbers, whereas `integer` reflects that the `MPG.city` column contains integers. 


### Hint 2: data.frame dimensions
To loop through the columns of the `data.frame`, you need to find out the dimensions. The total number of columns (or rows) can be obtained by [`ncol()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/nrow.html){target="_blank"} (or [`ncol()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/nrow.html){target="_blank"}).

### Hint 3: select columns based on a vector
To select multiple columns in a `data.frame` (if you forgot about column and row selection, see page 66 and beyond in [@Matloff2011]), you can use a vector with numbers:
```{r,collapse=T,output.lines=5}
# make a vector containing the column numbers that you want to select
# in this case: column 1, column 2, column 4
column.select <- c(1,2,4)
Cars93[,column.select]
```


## Nested for loops
In this part of the exercises, we will be looking at nested `for` loops

## Parameter combinations



```{r include=FALSE}
knitr::read_chunk("code/answers_other_loops.r")
```

```{r echo=FALSE, eval=TRUE, results=TRUE}
<<ex_param_combis>>
```
