# Looping with datasets

In this set of exercises, we will use existing
datasets and use looping techniques to read in the data. 
For example, let's say you have 10 datasets which you
need to read into R into one big dataframe. How to go
about this? Along the way you will encounter a bunch of functions from
various packages that are helpful when working with data files.

## Preliminaries
When generating large databases, we typically have to join together multiple files. Let us download and save a zip archive containing a bunch of Excel files from
[here](https://github.com/bramkuijper/stress/raw/master/figs/compare_dp/some_data.zip). Unzip the folder in some directory.

There are also some other files in the zip file, they should be maintained there for the sake of the exercise.

```{r, include=FALSE}
knitr::read_chunk("code/answers_loops_data.r")
```
## Obtaining a list of files
Use the function [`list.files()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/list.files.html) to return a list of files starting with `sheet_` and with the extension `.xls`. The list should return no other files, and hence you will have to make use of the `pattern` argument of the `list.files()` function to only list the desired files. The resulting output should look like the following:

```{r echo=FALSE, eval=TRUE, results=TRUE}
# output for exercise 3.1
<<ex_ch3-1>>
```
Hence, other files that are contained in the zip folder, like `another_excel_file.xls` or `sheet_other.txt`,  should not feature in the resulting list.

## Obtaining a list of files with full path names {#fileList}
Again, use the function [`list.files()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/list.files.html) from above, but make sure that it returns the complete path names, rather than just only the names of the sheets. Read the documentation of [`list.files()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/list.files.html) to see which argument(s) you will have to tweak in order for it to return full path names. Output should look like the following, except that your file system is different:
```{r echo=FALSE, eval=TRUE, results=TRUE}
<<ex_ch3-2>>
```

In future uses of `list.files()`, retaining the full path name saves a lot of hassle, as it means that any future script can use the output of `list.files()` without having the locate the actual directory of the excel sheets.

We store the list of files that you obtained using the `list.files()` function above for future use, in a variable `xls_sheet_list`. 

## Open a single file
Use the [`read_excel()`](https://readxl.tidyverse.org/reference/read_excel.html) function from the [readxl](https://readxl.tidyverse.org/) package to read the first excel file of the list (i.e., `xls_sheet_list[[1]]`) into a variable. Specifically, [`read_excel()`](https://readxl.tidyverse.org/reference/read_excel.html) reads data from the excel sheet into a so-called [`tibble`](https://tibble.tidyverse.org/), which is [`tidyverse`]( https://www.tidyverse.org/) 's more modern version of a [`data.frame`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/data.frame.html).

Subsequently, use `print()` to display the file's path name and then, below, print the first five lines of the resulting tibble. One can do so by using the [`slice_head()`](https://dplyr.tidyverse.org/reference/slice.html) function from the [`dplyr`](https://dplyr.tidyverse.org/) package. The resulting output should look like the following:
```{r echo=F, eval=TRUE, results=TRUE}
<<ex_ch3-3>>
```

## Getting information about each file within a `for`-loop {#forLoop}
Now we will use a `for`-loop to go through the list of files and read each file into a `tibble` using [`read_excel()`](https://readxl.tidyverse.org/reference/read_excel.html). Subsequently, we can then request some information about the current excel sheet by interrogating the resulting `tibble`. Here, we merely print the path name of each file (which you already have done previously), the number of rows contained in the `tibble` (using either [`nrow()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/nrow.html) or `tidyverse`'s [`tally()`](https://dplyr.tidyverse.org/reference/tally.html)) and the names of all the columns using the [`names()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/names.html) function. Hence, output should look like the following:
```{r echo=F, eval=TRUE, results=TRUE, R.options=list(max.print=5)}
<<ex_ch3-4>>
```

## Accumulating all `tibble`s into a single big `tibble`
Now try to accumulate all `tibble`s into a big one. Before we start our `for` loop, we assign an empty `tibble` (i.e., `tibble()`) to a variable, say `super_tbl`. Starting with an empty `tibble` makes sure that when we run our script again, `super_tbl` does not retain the contents of any previous runs of the same script.

Then, during each iteration of the `for`-loop, we append the contents of each excel sheet to this `super_tbl` variable by using the function [`bind_rows()`](https://dplyr.tidyverse.org/reference/bind.html).  We then assign the result of this function back to `super_tbl`, so that `super_tbl` grows during each iteration with the contents of each excel sheet.

Finally, we then just use  `print(super_tbl)` and the [`summary()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/summary.html) function to inspect the result
```{r echo=F, eval=TRUE, results=TRUE, R.options=list(max.print=5)}
<<ex_ch3-5>>
```


## Plotting the data
The next task is to use [`ggplot2`](https://ggplot2.tidyverse.org/) to plot the contents of `super_tbl`. Specifically, we plot the size on the $y$-axis and the different treatments on the $x$-axis, while the colors of the points should reflect the different individuals. We use points, rather than lines, by using [`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html). On [`ggplot2`](https://ggplot2.tidyverse.org/)'s website you can also download an elegant cheat sheet that is very helpful in achieving this.

```{r echo=F, eval=TRUE, results=TRUE, R.options=list(max.print=5)}
<<ex_ch3-6>>
```

## Data modification / conversion
As you can see when you `print()` the contents of `super_tbl`, all data types have the type of a `<dbl>` (double), which stands for double-precision floating point value. Such floating point values are typically used for data which varies in a continuous fashion, as is the case for the `size` column in `super_tbl`.

However, we often code treatments, and sometimes also individuals (if there are only a few, like here), as discrete factors at a nominal scale, meaning that number '1' is not necessarily lower than '2', it is merely an identifier of that treatment or individual. To accomplish this, we convert the `individual` and `treatment` column from `<dbl>` to `<fct>`, which stands for factor. 

To convert values into factors, we might either use the old-style function [`as.factor()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html) or we might use the more modern [`as_factor()`](https://haven.tidyverse.org/reference/as_factor.html) function from the [haven](https://haven.tidyverse.org/index.html) package. For this example there is no difference between both functions, but in general `as_factor()` is slightly more generic.

How to change our data using the `as_factor()` function? What we can do is apply the function to the columns in question, by using the [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html) function from the `dplyr()` package. Using this `mutate()` function (you have been checking the examples on the manual page, right?), we can then add two new columns to our data frame, say `individual_fac` and `treatment_fac` which reflect our new factor data.

If we now plot our new `individual_fac` and `treatment_fac` columns, our plot will look slightly different:

```{r echo=F, eval=TRUE, results=TRUE, R.options=list(max.print=5)}
<<ex_ch3-7>>
```

Can you describe what is different between this plot and the previous one and why?

## A further aside: regular expressions
The next exercise focuses on changing the data during the `for` loop that we used in section \@ref{forLoop}.

Consider the following scenario: your supervisor tells you (belatedly, as always) that they have omitted some relevant detail in the previous analysis. It turns out that the size measurements in the first 4 excel sheets (i.e., `sheet_1.xls`, `sheet_2.xls`, `sheet_3.xls` and `sheet_4.xls`) were performed on a different breed of individuals (breed A), while measurements from other excel sheets used animals of breed B. 

Let us visually inspect whether the relationship between treatment and size is different for the two different breeds of animals. To this end, we do not modify the underlying excel sheets, rather we add the column `breed` to the `tibble`s when reading in the data from each excel sheet. The `breed` column should have the values `A` for the first four sheets and the value `B` for the remaining sheets respectively.

A first suggestion may be to simply keep track of at which of the excel sheets we are when looping over the `xls_sheet_list` variable which contains all the files (see section \@ref{forLoop}), and then set the `breed` variable according to an `if` `else` statement (i.e., first 4 files `A`, remaining files `B`). However, if you look at the listing of files in section \@ref{fileList} above, you will see that `sheet_10.xls` (breed `B`) is at the second position in the variable `xls_sheet_list`, so using a simple rule like that is not going to work.

### Regular expressions to the rescue
Hence, we need to inspect the filenames themselves and sort out `sheet_1.xls`, `sheet_2.xls`, `sheet_3.xls` and `sheet_4.xls` from the rest. How to do this? [Regular expressions](https://stat.ethz.ch/R-manual/R-patched/library/base/html/regex.html) are a great way to filter numbers from textual data.

While we are not delving into regular expressions now, 
