--- 
title: "Various exercises on programming in R"
author: "Bram Kuijper"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: ["refsR.bib"]
biblio-style: apalike
link-citations: yes
description: "A brief set of exercises on control flow and loops in R"
---

# Introduction

Here a small selection of exercises to get to grips with R programming. It assumed you already have read about variables and simple control flow tools, such as `for`-loops and `if`-`else` statements. If you need more background on these topics, check out chapter 7 of [@Matloff2011] (I can borrow you a copy available), chapter 21 of [@Wickham2017] (available online) or check online for videos on these topics. For example, using the search terms `for loops in R` provides a bunch of great videos. 



## Hints 
Sometimes there are hints provided in the exercises, for example:
"You can do this by using the [`paste()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/paste.html) function", which involves
you reading the documentation on this function and trying
some of the provided examples in the documentation. I typically provide hyperlinks to the documentation, which is why the [`paste()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/paste.html) function here is given in blue.

## Some additional coding tips:

### Try out all examples & mistakes are good
When reading a book, try the code examples and snippets for yourself, by typing them out (don't copy-paste them, unless it is a massive amount of code). Only by typing out do you see what's going on and will you start to memorize the different coding constructs. Moreover, you are bound to make occasional mistakes! And mistakes are good, as you'll need to learn how to solve mistakes.

### Search enginges are your friend
* Search engines are your friend. If some function or construct is used that is new to one, google what it is about. Say, one encounters the statement 
```{r echo=TRUE, results=TRUE}
means <- apply(X=cars,MARGIN=2,FUN=mean)
```
and one does not know what the [`apply`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html) statement is about or what the `cars` variable is about, I'd immediately search for `apply r` or `cars r`. 

### Comment your code
Add comments to your code, by starting a line with a hashtag character `#`. For example:
```{r echo=TRUE, results=TRUE}
# list of TRUE / FALSE VALUES dependent on whether number is even
even_odd <- c(1,2,3,4,5,6,7,8) %% 2 == 0
```

### Keep an error record
Try to keep a record of all the main error messages you encounter, so that you build a record of different error messages and how they have eventually been resolved. This serves as a great lookup tool for reccurring errors

### Use `traceback()`
If you stumble upon an error, use R's [`traceback()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/traceback.html) function to find the root cause.

### Inspect values using `print()` or use R's debugger
When you have just started to code in R, use as many [`print()`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/print.html) statements as possible, to obtain information about the state of your program and the values of your variables. When you become more experienced, you can start to inform yourself about R's very handy debugging tools that can do all the variable inspection work for you. Read more about R debugging in the freely available [chapter 22](https://adv-r.hadley.nz/debugging.html) of [@Wickham2019].
